---
title: "Analyzing Voting Difficulty"
author: "Sonia Song, Kenneth Hahn, Mei Qu"
subtitle: "Lab 1: Datasci 203" 
output:
  pdf_document:
    toc: yes
    fig_caption: yes
  bookdown::pdf_document2:
    number_sections: yes
    toc: yes
    output:
---

```{=tex}
\newpage
\setcounter{page}{1}
```
```{r Load packages and set options, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyverse)
library(tidyr)
library(knitr)
library(kableExtra)
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(echo=FALSE, message=FALSE)
```

```{r Load data, include=FALSE}
raw_data <- read_csv("./data/anes_pilot_2022_csv_20221214.csv")
nrow(data)
data <- raw_data[!is.na(raw_data$weight) & raw_data$weight != "", ]
nrow(data)
```

# Importance and Context

Voter turnouts in U.S. elections have historically been below two thirds of the eligible voting population. With about 66% of the eligible voting population turnout, 2020 presidential election saw one of the highest rate for any national election since 1900.\footnote{Hartig, H., Daniller, A., Keeter, S., \& Green, T. V. (2023, July 12). 1. Voter turnout, 2018-2022. Pew Research Center. https://www.pewresearch.org/politics/2023/07/12/voter-turnout-2018-2022/}

Given the rising political divide in the U.S., any voting irregularities can potentially create an outsized effect on election outcomes. One of those variables is difficulty of voting. Our analysis seeks to answer the below research question using statistical methods:

```{=tex}
\begin{quote}
  \textit{Do Democratic voters or Republican voters experience more difficulty voting? }
\end{quote}
```
The answer to this question can provide additional transparency, increasing public confidence in elections. Moreover, further investigation of the underlying drivers of voting difficulty can provide valuable insights for improving political and civic engagement. 

```{r Classify voters, include=FALSE}
data <- data %>% filter((reg == 1 | reg == 2) | votehard != -1)
nonVoters = data %>% filter(reg == 3 & votehard == -1)
nrow(data)
nrow(nonVoters)
```

```{r Classify voters by democrat or republican, include=FALSE}
# Democrat if they identified as democrat or lean towards democrat
dem <- data$pid1d == 1 |  data$pid1r == 1 |  data$pidlean == 2
rep <- data$pid1d == 2 |  data$pid1r == 2 |  data$pidlean == 1
data$party <- if_else(dem, 'Democrat',
                      if_else(rep, 'Republican',NA))

#Check number of data points in each category
print(paste('Number of Democrat Voters: ', sum(na.omit(data$party == 'Democrat'))))
print(paste('Number of Republican Voters: ', sum(na.omit(data$party == 'Republican'))))
print(paste('Number of Independent/Other Voters: ',sum(is.na(data$party))))
```

# Data and Methodology

Our analysis uses the American National Election Studies (ANES) 2022 Pilot Study dataset. This is an observational dataset based on sample respondents collected from YouGov. There are a total of `r nrow(raw_data)` cases in the study. We removed 85 unweighted cases not selected by the sample matching procedure, which the study recommends excluding for making any inferences about the general population.

We define voters as those who are registered to vote (responded 1 or 2 to `reg`) or those that answered the "how difficult was it to vote" question (response to `votehard` was not equal to -1, or a skipped answer). We used an "or" statement because there could be voters who are not currently registered to vote but voted in the November 8th election and they should be accounted for in our analysis. We note from the documentation that only respondents who "definitely voted" or "probably voted" received the `votehard` question. By defining voters as such, we realize we may miss responders who voted in previous elections if they did not register to vote and skipped the `votehard` question. However, we observed that there were no respondents in this category, making our definition holistic.

To differentiate a Democrat from a Republican voter, we recognize that the survey generates a `rand_pid` (a random integer value) from 1-3 for each respondent. Respondents who get assigned a `rand_pid` of 1 or 3 receive question `pid1d` and respondents who get assigned a `rand_pid` of 2 receive `pid1r`. Both ask the question "Generally speaking, do you usually think of yourself as a Democrat, a Republican, an independent?" but with different phrasing of the questions and answers. We also see a question `pidlean` asking respondents which party they are closer to if they stated they were "Independent", "Something else", or skipped the earlier question. Therefore, we categorized Democrats as those who responded "Democrat" to question `pid1d`, `pid1r` or `pidlean` and Republicans as those who responded "Republican" to those same questions, resulting in `r nrow(data)` responses. We considered using the variable `pid_x` which is a ranking from 1-7 from "Strong Democrat" to "Strong Republican", but we chose not to use this variable as it was not mentioned in the Questionnaire Specifications. We also decided not to further categorize respondents who had not been categorized to avoid false assumptions on their political views.

```{r Determine difficulty voting, include=FALSE}
# divide the data into 2 subgroups by democrat and republican and filter for data with only Democrat and Republican

party_data <- data[!is.na(data$party) & data$votehard != -1,]
#Generate data for the test, removing legitimate skip "-1" as it does not indicate whether or not the vote experiences any difficulty
dem_data_votehard <- subset(party_data, party_data$party == "Democrat")
rep_data_votehard <- subset(party_data, party_data$party == "Republican")
```

The survey asks the respondents “How difficult was it for you to vote?” (also known as `votehard`) with 5 choices ranging from “Not difficult at all” (1) to “Extremely difficult” (5). Once we filtered the data for Democratic and Republican voters, we used the `votehard` ranking to conduct our statistical test to test the null hypothesis, as this variable is directly applicable to the research question at hand. After removing anyone who is not a Democrat or Republican and removing those who did not answer the `votehard` question ( `votehard` = -1), we remain with `r nrow(party_data)` rows. We also considered using the question “How difficult was it for you to register to vote?” for inclusion of voters with difficulty registering but decided to leave it out because only 12% in the eligible sample did not skip the question and based on the questionnaire those who responded were voters who successfully registered to vote, so the responses are not sufficiently informative.

The sample test we choose is the Wilcoxon rank-sum test. The data is not paired as each respondent has their own individual score. We believe that this is the most appropriate test as the dataset satisfies the following assumptions for the Wilcoxon rank-sum test: 1. The data is I.I.D. because one respondent's answer does not depend on the other and both groups are pulled from the same distribution 2. Data is ordinal, not metric, because the difference between intervals, such as a 4 (very difficult) to a 5 (extremely difficult), is not universally agreed upon. Because the data is ordinal, it must also be non-parametric. Given that the data is at ordinal and I.I.D., we must use the Hypothesis of Comparisons to define our null hypothesis, where P is the probability, and we are comparing the `votehard` scores for Democrats and Republicans, respectively:

```{=tex}

\textbf{$H_0$}: $P(votehard_{Democrat}>votehard_{Republican})=P(votehard_{Democrate}<votehard_{Republican})$

```

We will be utilizing a two-tailed Wilcoxon Rank-Sum test because a one tailed test will not only make it easier to reject the null hypothesis but also assumes that the opposite case cannot occur, which in this scenario either probability can be just as likely to occur. We will also be defining our type I error as $\alpha = 0.05$ as our threshold for rejecting the null hypothesis. As a result our alternative hypothesis is as follows:

\textbf{$H_a$}: $P(votehard_{Democrat} > votehard_{Republican}) \neq P(votehard_{Democrat} < votehard_{Republican})$

# Results

```{r Get breakdown of voters and their votehard score, echo = FALSE}

overall_summary_by_party <- party_data[,c('party','votehard')] %>%
  group_by(party) %>%
  summarise(
    Count_votehard = n(),
    Mean_votehard = mean(votehard),
    Standard_Deviation_votehard = sd(votehard)
  )

kable(overall_summary_by_party,format='latex',booktabs=TRUE,caption='Votehard Summary Table by Party')%>%
  kable_styling(latex_options = c("striped", "hold_position"))


```

Table 1 above shows the results of filtering our dataset for only Democratic and Republican voters. The table portrays that there are `r Reduce("-",overall_summary_by_party$Count_votehard)` more Democrat voters than Republican voters and that Democrats have an average `votehard` score that is `r round(Reduce("-",overall_summary_by_party$Mean_votehard)*100,2)`% higher than Republicans. Because this data is ordinal, we cannot directly compare the means of the two groups and we conducted the Wilcoxon Rank-Sum test with the code below.

```{r Wilcoxon rank-sum test, echo=TRUE}
results <- wilcox.test(dem_data_votehard$votehard, 
                       rep_data_votehard$votehard, alternative = "two.sided")
```

After performing the Wilcoxon rank-sum test, we observed that our p-value is less than 0.05. This means that the probability that a Democrat has more difficulty voting than a Republican is not equal to the probability that a Democrat has less difficulty voting than a Republican (W = `r results$statistic`, p = `r results$p.value`). Figure 1 delineates the differences in responses between the two survey groups at each difficulty level. We calculated by the percentage of the respective party that answered each of the categories for votehard, since there are more Democrat voters who answered the question than Republican voters (`r nrow(dem_data_votehard)` Democrat voters and `r nrow(rep_data_votehard)` Republican voters).

```{r Determine Difficulty Bar Chart Calculations, include=FALSE}
# get only the party and votehard scores
party_data_difficulty <- party_data[,c('votehard','party')]

# calculate the total count in each party
total_count <- party_data_difficulty %>%
  group_by(party) %>%
  summarise(
    count = n()
  )

n_democrats <- as.integer(total_count[total_count$party == 'Democrat','count'])
n_republicans <- as.integer(total_count[total_count$party == 'Republican','count'])

# look at summary stats of party data 
# use total_count to create a new column percentage_score that gets the number of count_votehard and divides by the total_count of each respective party
# percentage_score is calculated to remove the fact that there are more democrats than republicans in this dataset.
summary_by_party <- party_data_difficulty %>%
  group_by(party,votehard) %>%
  summarise(
    mean_votehard = mean(votehard),
    median_votehard = median(votehard),
    count_votehard = n(),
    std_votehard = sd(votehard),
    .groups = 'drop_last'
  ) %>%
  left_join(total_count, by = 'party') %>%
  mutate(percentage_score=count_votehard/count * 100)

percent_democrats_not_difficult <- as.integer(summary_by_party[(summary_by_party$party == 'Democrat') & (summary_by_party$votehard == 1),'percentage_score'])
percent_republicans_not_difficult <- as.integer(summary_by_party[(summary_by_party$party == 'Republican') & (summary_by_party$votehard == 1),'percentage_score'])

print(summary_by_party)

summary_by_party$votehard_string <- as.character(summary_by_party$votehard)

votehard_label <- c(
  '1' = 'Not difficult at all',
  '2' = 'A little difficult',
  '3' = 'Moderately difficult',
  '4' = 'Very difficult',
  '5' = 'Extremely difficult'
)
```

```{r Determine Difficulty Bar Chart, fig.cap='Percentage of Respective Party that Answered \"votehard\" from 1-5', fig.width=8, fig.height=2}
# Creates a barplot that compares the two parties based on the percentage of each respective party and what score they voted for from 1 to 5.
ggplot(summary_by_party) + 
  aes(x = votehard_string, y = percentage_score, fill = party) +
  geom_bar(stat = 'identity', position = 'dodge') +
  labs(x = 'Difficulty Score', y = 'Percentage of Party (%)') +
  scale_fill_manual(values = c('Democrat' = 'blue', 'Republican' = 'red')) +
  scale_x_discrete(labels = votehard_label)
```

Figure 1 shows a heavy tail where a vast majority of both Democrats and Republicans did not think that voting was difficult at all (`r percent_democrats_not_difficult`% and `r percent_republicans_not_difficult`%, respectively). We also do observe that a higher percent of Democrats responded to the “more difficult” option than Republicans did. Ultimately, the Wilcoxon Rank-Sum test concludes that we can reject our null hypothesis that the probabilities are not equal; however, the limitations of the test cannot conclude which probability is more likely to occur and the result is restricted to the sample set of respondents, meaning that we cannot apply these results directly to the U.S. population.

# Discussion
The study found evidence that the probability that Democratic voters find voting more difficult than Republican voters is not equal to the probability that Republican voters find voting more difficult than Democratic voters. Our results may be of key interest to future political campaigns, who may have the goal of increasing voter turnout for their party. While this study addresses overall voting difficulty, future studies may build on this framework to further understand the implications behind voter:

1. Conduct more statistical tests (e.g. a two-proportion test) at the category level. For example, we can utilize the 10 different `vharder` categories, which is a series of questions where respondents can select whether they found a specific reason that made it difficult to vote or not. Upon initial investigation with Figure 2, we observed that for each of the categories, a higher percentage of Democratic voters responded than Republican voters.  Make better inferences about the population by utilizing the weights that ANES provides while conducting the sample test.

2. Make better inferences about the population utilizing the weights that ANES provides while conducting the sample test 


```{r Plot Reasons for Difficulty by Party Bar Chart Calculations, include=FALSE}

# creates a long form pivot table to get each vharder column into separate rows
party_data_long <- party_data %>% 
  pivot_longer(cols = c('vharder_0','vharder_1','vharder_2','vharder_3','vharder_4',
                        'vharder_5','vharder_6','vharder_7','vharder_8','vharder_9',
                        'vharder_10'), 
               names_to = "vharder",values_to = "vharder_response")

# only give the data where someone responded to a vharder question
vharder_filtered <- party_data_long %>%
  filter(vharder_response == 1)

# count the number of people who voted for each of the vharder categories and then calculate the percentage of the party that selected that vharder option using the total_count calculated above
count_vharder <- vharder_filtered %>%
  group_by(vharder,party) %>%
  summarise(count_vharder = n(), .groups = 'drop_last') %>%
  left_join(total_count,by='party') %>%
  mutate(percentage_vharder=count_vharder/count * 100)

# create labels to classify the vharder categories
vharder_label <- c(
  'vharder_0' = 'Registration',
  'vharder_1' = 'Wait Time',
  'vharder_2' = 'Identification Card',
  'vharder_3' = 'Voting Machine',
  'vharder_4' = 'Polling Place Location',
  'vharder_5' = 'Commute to Polling Place',
  'vharder_6' = 'Transportation Cost',
  'vharder_7' = 'Work Schedule',
  'vharder_8' = 'Weather',
  'vharder_9' = 'Absentee Ballot',
  'vharder_10' = 'No Postage'
)
```

```{r Plot Reasons for Difficulty by Party Bar Chart, fig.cap='Reason for Difficulty Voting by Political Party', fig.width=8, fig.height=3}
# create a bar plot for each of the reasons and split by party
# using a percentage of the party because the number of democrats is greater than the number of republican
ggplot(count_vharder) +
  aes(x=vharder, y=percentage_vharder, fill=party) +
  geom_bar(stat='identity',position='dodge') +
  labs(x='Reason for Difficulty Voting',y='Percentage of Party (%)') +
  scale_fill_manual(values = c('Democrat' = 'blue','Republican'='red')) +
  theme(axis.text.x = element_text(angle = 30,vjust=1,hjust=1)) +
  scale_x_discrete(labels = vharder_label)
```
